jax-cosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-01-22 paper.tex
Updates from Overleaf
SVI+NeutraReparam start

@@

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

>>>>

%%%% new command (JEC)
\newcommand{\nn}{\nonumber}
\newcommand{\numpyro}{\texttt{NumPyro}}
\newcommand{\bydef}{\overset{def.}{\equiv}}

@@

%
\subsection{Use of SVI \& Neutral Reparametrisation for NUTS}
%
Let us remind briefly what is the \textit{Stochastic Variational Inference} method \citep{2012arXiv1206.7051H, 2017arXiv171105597Z} and show where auto-differentiation helps. If one  writes $p(z)$ the prior, $p(\mathcal{D}|z)$ the likelihood and $p(\mathcal{D})$ the marginal likelihood, then thanks to the  Bayes theorem one gets $p(z|\mathcal{D})=p(z)p(\mathcal{D}|z)/p(\mathcal{D})$ the posterior distribution of a model with latent variables $z$ and a set of observations $\mathcal{D}$. The SVI aims to get an approximation of this distribution, ie. $p(z|\mathcal{D}) \approx q(z;\lambda)$, by determining the variational parameters $\lambda$ of a predefined distribution. To do so, one uses the Kullback-Leibler divergence of the two distributions $KL(q(z;\lambda)||p(z|\mathcal{D}))$ leading to the following relation
\begin{align}
\log p(\mathcal{D}) &= \mathtt{ELBO} +  KL(q(z;\lambda)||p(z|\mathcal{D})) \label{eq-ELBO} \\
\mathrm{with} \ \mathtt{ELBO} &\equiv -\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right] + \mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right] 
\end{align}
which defines the evidence lower bound (ELBO) that one tends to maximize to get the $\lambda$ values and the  highest lower bound on the logarithm of the marginal probability of the observations $\log p(\mathcal{D})$. So, the optimal variational distribution satisfies
\begin{equation}
q(z;\lambda^\ast) = \underset{q(z;\lambda)}{\mathrm{argmax}}\  \mathtt{ELBO} = 
\underset{\lambda}{\mathrm{argmin}}\ \mathcal{L}(\lambda)
\end{equation}
The function $\mathcal{L}(\lambda)$ is the cost function used in practice. It is composed of two parts:
\begin{equation}
\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log q(z;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(z;\lambda)}\left[ \log p(z,\mathcal{D}) \right]}_{model}
\label{eq-loss-svi-1}
\end{equation}
where the \textit{guide}, in the \numpyro\ word, may be a multi-variate gaussian distribution (MVN) for instance.

Using the auto-differentation tool, one can use "black-box" guides (aka \textit{automatic differentiation variational inference}). As stated by the authors of \citep{2016arXiv160300788K} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
function, taking derivatives, and running a gradient-based or coordinate-ascent optimization. First is defined a bijective differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$. One can develop "AutoGuides" (\numpyro\ terminology) that can be adapted to the user models. Then, the cost function reads
\begin{equation}
\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log q(\xi;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log p(\xi,\mathcal{D}) \right]}_{model}
\label{eq-loss-svi-2}
\end{equation}
with
\begin{equation}
p(\xi,\mathcal{D}) \bydef p(T^{-1}(\xi),\mathcal{D}) |J_{T^{-1}}(\xi)|
\end{equation}
which exhibit the Jacobian of the $T^{-1}$ transformation. The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints.

Now, one is faced with the evaluation of the expectations during gradient descent. The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}). Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ ($\lambda=(\mu, \Sigma)$). One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg. $\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$. The Jacobien of this distribution is 1, so the cost function reads
\begin{multline}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\eta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
\label{eq-loss-svi-3}
\end{multline}
where $\mathbb{H}(q)$ is the Shannon entropy of the gaussian distribution which is independent of the model, so its gradient can be computed once for all and reused in user model. Then,
to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to
\begin{multline}
-\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
+ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
\right\}
+ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
\label{eq-loss-svi-4}
\end{multline}
For the use case of the MVN (auto)guide, one gets
\begin{align}
-\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
+ \nabla_\xi \log|J_{T^{-1}}(\xi)|
\right\} \nn \\
-\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
\right\}+ (L^{-1})^T
\end{align}
Reminding that $z=T^{-1}(\xi)$ and $\xi=S^{-1}(\zeta)$, the expectation used in the above expression is computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$.
%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix
